Weighted Model Counting with Conditional Weights for Bayesian Networks

### 2021-05-28 doc/UAI_paper/paper.tex
The poster is halfway decent (and some other small changes)

\author[1]{\href{mailto:Paulius Dilkas <p.dilkas@sms.ed.ac.uk>?Subject=Your UAI
    2021 paper}{Paulius~Dilkas}{}}
\author[1,2]{Vaishak~Belle}
\affil[1]{%
    School of Informatics\\
    University of Edinburgh\\
    Edinburgh, UK
}
\affil[2]{%
    Alan Turing Institute\\
    London, UK
}

@@

While in this paper we focus on measure-theoretic foundations and propose a
new encoding for Bayesian networks, a related line of work
\citep{my_sat_paper}---motivated by the same issue of WMC encodings having both
more variables and more clauses as a consequence of having to conform to an
unnecessarily restrictive format---shows how these variables and clauses can be
removed from (most) already-existing Bayesian network encodings. Although tested
on Bayesian networks, this variable removal is formulated in a way that applies
to arbitrary WMC instances.
% I could also mention DPMC vs ADDMC and pseudo-Boolean functions

@@

\section{Experimental Comparison} \label{sec:experiments}

@@

While we chose \textsf{ADDMC} \citep{DBLP:conf/aaai/DudekPV20} as the WMC
algorithm and Bayesian networks as a canonical example of a probabilistic
inference task, these are only examples meant to illustrate the broader idea
that choosing a more expressive representation of weights can outperform
increasing the size of the problem to keep the weights simple. Indeed, in this
work, we have provided a new theoretical perspective on the expressive power of
WMC and illustrated the empirical benefits of that perspective. The same idea
can be adapted to other inference problem domains such as probabilistic programs
\citep{DBLP:journals/tplp/FierensBRSGTJR15,DBLP:journals/corr/abs-2005-09089} as
well as to search-based solvers such as \textsf{Cachet}
\citep{DBLP:conf/sat/SangBBKP04} and \textsf{DPMC} ---an extension to
\textsf{ADDMC} that adds support for computations based on tensors (rather than
ADDs) and planning based on tree decompositions \cite{DBLP:conf/cp/DudekPV20}.
Another important direction for future work is to develop a better understanding
of what properties of Bayesian networks make an instance easy for some
algorithm-encoding combinations more than others.

>>>>

\author{Paulius~Dilkas}
\author{Vaishak~Belle}
\affil{%
    University of Edinburgh\\
    Edinburgh, UK \\
    \texttt{p.dilkas@sms.ed.ac.uk}, \texttt{vaishak@ed.ac.uk}
}

@@

While in this paper we focus on measure-theoretic foundations and propose a
new encoding for Bayesian networks, a related line of work
\citep{my_sat_paper}---motivated by the same issue of WMC encodings having both
more variables and more clauses as a consequence of having to conform to an
unnecessarily restrictive format---shows how these variables and clauses can be
removed from (most) already-existing Bayesian network encodings.
% I could also mention DPMC vs ADDMC and pseudo-Boolean functions

@@

\section{Experimental Results} \label{sec:experiments}

@@

While we chose \textsf{ADDMC} \citep{DBLP:conf/aaai/DudekPV20} as the WMC
algorithm and Bayesian networks as a canonical example of a probabilistic
inference task, these are only examples meant to illustrate the broader idea
that choosing a more expressive representation of weights can outperform
increasing the size of the problem to keep the weights simple. Indeed, in this
work, we have provided a new theoretical perspective on the expressive power of
WMC and illustrated the empirical benefits of that perspective. Perhaps the same
idea could be adapted to other inference problem domains such as probabilistic
programs
\citep{DBLP:journals/tplp/FierensBRSGTJR15,DBLP:journals/corr/abs-2005-09089} as
well as to search-based solvers such as \textsf{Cachet}
\citep{DBLP:conf/sat/SangBBKP04} and \textsf{DPMC} ---an extension to
\textsf{ADDMC} that adds support for computations based on tensors (rather than
ADDs) and planning based on tree decompositions \cite{DBLP:conf/cp/DudekPV20}.

### 2021-05-28 doc/poster/poster.tex
The poster is halfway decent (and some other small changes)

\usepackage{booktabs}
\usepackage{mathtools}
\usepackage{tikz-cd}
\usepackage{authblk}
\usepackage[export]{adjustbox}

%\addbibresource{../UAI_paper/paper.bib}

@@

% TODO: use lots of bold

@@

\makeatletter
\def\maketitle{\AB@maketitle}
\makeatother

\title{Weighted Model Counting with Conditional Weights for Bayesian Networks}
\author[1]{Paulius Dilkas}
\author[1,2]{Vaishak Belle}
\affil[1]{School of Informatics, University of Edinburgh, Edinburgh, UK}
\affil[2]{The Alan Turing Institute, London, UK}
\titlegraphic{\includegraphics[height=200pt,valign=c]{logo_inf.png}\hspace{3em}\includegraphics[height=200pt,valign=c]{logo_ecr.png}\hspace{3em}\includegraphics[height=100pt,valign=c]{logo_ukri.png}}

@@

\begin{columns}
  \column{0.5}
  \block{Boolean Algebras and Propositional Logic}{
    Let $U = \{ a, b \}$. Then $2^{2^U}$ is a Boolean algebra with the following
    Hasse diagram ($x \le y$ if $x \subseteq y$ or, equivalently, $x = x \land
    y$).
    % \[
    %   \begin{tikzcd}[ampersand replacement=\&, column sep=tiny]
    %     \& \& \& \& \& \top \ar[dlll,dash,gray] \ar[dl,dash,gray]
    %     \ar[dr,dash,gray] \ar[drrr,dash,gray] \& \& \& \\
    %     \& \& a \lor b \& \& b \to a \& \& a \to b \& \& \neg a \lor \neg
    %     b \\
    %     a \ar[urr,dash,gray] \ar[urrrr,dash,gray] \& \& b \ar[u,dash,gray]
    %     \ar[urrrr,dash,gray] \& \& a \leftrightarrow b \ar[u,dash,gray]
    %     \ar[urr,dash,gray] \& \& a + b \ar[ullll,dash,gray] \ar[urr,dash,gray]
    %     \& \& \neg b \ar[ullll,dash,gray] \ar[u,dash,gray] \& \& \neg a
    %     \ar[ullll,dash,gray] \ar[ull,dash,gray] \\
    %     \& \& \fbox{$a \land b$} \ar[ull,dash,gray] \ar[u,dash,gray]
    %     \ar[urr,dash,gray] \& \& \fbox{$a \land \neg b$} \ar[ullll,dash,gray]
    %     \ar[urr,dash,gray] \ar[urrrr,dash,gray] \& \& \fbox{$\neg a \land b$}
    %     \ar[ullll,dash,gray] \ar[u,dash,gray] \ar[urrrr,dash,gray] \& \&
    %     \fbox{$\neg a \land \neg b$} \ar[ullll,dash,gray] \ar[u,dash,gray]
    %     \ar[urr,dash,gray] \\
    %     \& \& \& \& \& \bot \ar[ulll,dash,gray] \ar[ul,dash,gray]
    %     \ar[ur,dash,gray] \ar[urrr,dash,gray] \& \& \&
    %   \end{tikzcd}
    % \]
  }

  \block{Experimental Results}{
    \begin{tikzfigure}[Cumulative numbers of instances solved by combinations of
      algorithms and encodings over time.]
      \centering
      \includegraphics[width=\linewidth]{cumulative}
    \end{tikzfigure}
    \begin{tikzfigure}[An instance-by-instance comparison between
      $\textsf{ADDMC} + \texttt{cw}$ and the best overall combination of
      algorithm and encoding ($\textsf{Ace} + \texttt{cd06}$, on the left) as
      well as the second-best encoding for \textsf{ADDMC} (\texttt{sbk05}, on
      the right).]
      \centering
      \includegraphics[width=\linewidth]{scatter}
    \end{tikzfigure}
  }

@@

  \block{TODO}{
    \begin{tikztable}[Asymptotic upper bounds on the numbers of variables and
      clauses/ADDs for each encoding.]
      \centering
      \begin{tabular}{l c c}
        \toprule
        Encoding(s) & Variables & Clauses/ADDs \\
        \midrule
        \texttt{bklm16}, \texttt{cd05}, \texttt{cd06}, \texttt{sbk05} & $O(nv^{d+1})$ & $O(nv^{d+1})$ \\
        \texttt{cw} & $O(nv)$ & $O(nv^2)$ \\
        \texttt{d02} & $O(nv^{d+1})$ & $O(ndv^{d+1})$ \\
        \bottomrule
      \end{tabular}
    \end{tikztable}
  }

>>>>

\usepackage{booktabs}
\usepackage{mathtools}
\usepackage{tikz-cd}
\usepackage[export]{adjustbox}
\usepackage{amsfonts}
\usepackage[capitalise]{cleveref}
\usepackage{subcaption}

@@

\usetikzlibrary{arrows}
\usetikzlibrary{arrows.meta}

\AtBeginEnvironment{tikzfigure}{\captionsetup{type=figure}}

@@

%\makeatletter
%\def\maketitle{\AB@maketitle}
%\makeatother

\title{Weighted Model Counting with Conditional Weights for Bayesian Networks}
\author{Paulius Dilkas, Vaishak Belle}
\institute{University of Edinburgh, Edinburgh, UK\\
  \texttt{p.dilkas@sms.ed.ac.uk}, \texttt{vaishak@ed.ac.uk}}
\titlegraphic{\includegraphics[height=200pt,valign=c]{logo_inf.png}\hspace{3em}\includegraphics[height=200pt,valign=c]{logo_ecr.png}\hspace{3em}\includegraphics[height=100pt,valign=c]{logo_ukri.png}}

@@

\begin{columns}

@@

  \block{Abstract}{
    \textbf{Weighted model counting (WMC)} has emerged as the unifying inference
    mechanism across many (probabilistic) domains. Encoding an inference problem
    as an instance of WMC typically \textbf{necessitates adding extra literals
      and clauses}. This is partly so because the predominant definition of WMC
    assigns weights to models based on weights on literals, and this severely
    restricts what probability distributions can be represented. We develop a
    \textbf{measure-theoretic perspective} on WMC and propose a way to encode
    conditional weights on literals analogously to conditional probabilities.
    This representation can be as succinct as standard WMC with weights on
    literals but can also expand as needed to represent probability
    distributions with less structure. To demonstrate the performance benefits
    of conditional weights over the addition of extra literals, we develop a
    \textbf{new WMC encoding for Bayesian networks} \texttt{cw} and adapt a
    state-of-the-art WMC algorithm \textsf{ADDMC} to the new format. Our
    experiments show that the new encoding significantly improves the
    performance of the algorithm on most benchmark instances.
  }

  \block{WMC as a Measure on a Boolean Algebra}{
    Let $U$ be a set. Then $2^{2^U}$ (the powerset of the powerset of $U$)
    is the Boolean algebra that directly corresponds to all possible
    propositional formulae constructable from $a$ and $b$. A \textbf{measure} on
    $2^{2^U}$ is a function $\mu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ such that
    \begin{itemize}
    \item $\mu(\bot) = 0$, and
    \item $\mu(a \lor b) = \mu(a) + \mu(b)$ for all $a, b \in 2^{2^U}$ whenever
      $a \land b = \bot$.
    \end{itemize}
    A \textbf{weight function} is a function $\nu\colon 2^U \to \mathbb{R}_{\ge
      0}$. A weight function is \textbf{factored} if
    \[
      \nu = \prod_{x \in U} \nu_x
    \]
    for some functions $\nu_x\colon 2^{\{x\}} \to \mathbb{R}_{\ge 0}$, $x \in
    U$. We say that a weight function $\nu\colon 2^U \to \mathbb{R}_{\ge 0}$
    \textbf{induces} a measure $\mu_\nu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ if
    \[
      \mu_\nu(x) = \sum_{\{u\} \le x} \nu(u).
    \]
    Finally, a measure $\mu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ is
    \textbf{factorable} if there exists a factored weight function $\nu\colon 2^U
    \to \mathbb{R}_{\ge 0}$ that induces $\mu$. In this formulation, WMC
    corresponds to the process of calculating the value of $\mu_\nu(x)$ for some
    $x \in 2^{2^U}$ with a given definition of $\nu$.
    \innerblock{Relation to the Classical (Logic-based) View of WMC}{
      Let $\mathcal{L}$ be a propositional logic with two atoms $a$ and $b$ and
      $w\colon \{ a, b, \neg a, \neg b \} \to \mathbb{R}_{\ge 0}$ a weight
      function defined as $w(a) = 0.3$, $w(\neg a) = 0.7$, $w(b) = 0.2$, $w(\neg
      b) = 0.8$. Furthermore, let $\Delta$ be a theory in $\mathcal{L}$ with a
      sole axiom $a$. Then $\Delta$ has two models: $\{ a, b \}$ and $\{ a, \neg
      b \}$ and its WMC is
      \begin{equation} \label{eq:wmc_example}
        \mathrm{WMC}(\Delta) = \sum_{\omega \models \Delta} \prod_{\omega \models l} w(l) = w(a)w(b) + w(a)w(\neg b) = 0.3.
      \end{equation}
      Alternatively, we can define $\nu_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge
        0}$ as $\nu_a(\{ a \}) = 0.3$, $\nu_a(\emptyset) = 0.7$ and $\nu_b\colon
      2^{\{b\}} \to \mathbb{R}_{\ge 0}$ as $\nu_b(\{ b \}) = 0.2$,
      $\nu_b(\emptyset) = 0.8$. Let $\mu$ be the measure on $2^{2^U}$ induced by
      $\nu = \nu_a \cdot \nu_b$. Then, equivalently to \cref{eq:wmc_example}, we
      can write
      \begin{align*}
        \mu(a) &= \nu(\{ a, b \}) + \nu(\{ a \}) \\
               &= \nu_a(\{a\})\nu_b(\{b\}) + \nu_a(\{a\})\nu_b(\emptyset) = 0.3.
      \end{align*}
      Thus, one can equivalently think of WMC as summing over \textbf{models of
        a theory} or over \textbf{atoms below an element} of a Boolean algebra.
    }
  }

  \block{Example: Encoding Bayesian Networks}{
    \begin{tikzfigure}[A Bayesian network with its conditional probability
      tables]
      \begin{subfigure}{0.2\linewidth}
        \centering
        \begin{tikzpicture}[edge from parent/.style={draw,-latex}, level distance=100pt, edge from parent/.style={draw,-{Latex[length=5mm]}}]
          \node[draw,circle] {$A$}
          child {node[draw,circle] {$B$}};
        \end{tikzpicture}
      \end{subfigure}%
      \begin{subfigure}{0.8\linewidth}
        \centering
        \begin{tabular}[t]{cc}
          \toprule
          $a$ & $\Pr(A = a)$ \\
          \midrule
          1 & 0.5 \\
          0 & 0.5 \\
          \bottomrule
        \end{tabular}
        \begin{tabular}[t]{ccc}
          \toprule
          $a$ & $b$ & $\Pr(B = b \mid A = a)$ \\
          \midrule
          1 & 1 & 0.6 \\
          1 & 0 & 0.4 \\
          0 & 1 & 0.1 \\
          0 & 0 & 0.9 \\
          \bottomrule
        \end{tabular}
      \end{subfigure}
    \end{tikzfigure}
    Let $U = \{ \lambda_{A=1}, \lambda_{B=1} \}$. The weight function $\nu\colon
    2^U \to \mathbb{R}_{\ge 0}$ for this network can be defined as $\nu
    \coloneqq \nu_A \cdot \nu_B$, where $\nu_A = 0.5$, and
    \begin{align*}
      \nu_B &= 0.6[\lambda_{B=1}] \cdot [\lambda_{A=1}] + 0.4\overline{[\lambda_{B=1}]} \cdot [\lambda_{A=1}] \\
            &+ 0.1[\lambda_{B=1}] \cdot \overline{[\lambda_{A=1}]} + 0.9\overline{[\lambda_{B=1}]} \cdot \overline{[\lambda_{A=1}]}.
    \end{align*}
  }

  \block{Experimental Results}{
    \begin{tikzfigure}[Cumulative numbers of instances solved by combinations of
      algorithms and encodings over time]
      \centering
      \includegraphics[width=\linewidth]{cumulative}
    \end{tikzfigure}
    \begin{tikzfigure}[An instance-by-instance comparison between
      $\textsf{ADDMC} + \texttt{cw}$ and the best overall combination of
      algorithm and encoding ($\textsf{Ace} + \texttt{cd06}$, on the left) as
      well as the second-best encoding for \textsf{ADDMC} (\texttt{sbk05}, on
      the right)]
      \centering
      \includegraphics[width=\linewidth]{scatter}
    \end{tikzfigure}
    We compare all six WMC encodings for Bayesian networks when run with both
    \textsf{ADDMC} and the WMC algorithms used in the original papers. The
    cumulative plot shows that \textsf{ADDMC} significantly \textbf{underperforms} when
    combined with any of the \textbf{previous encodings}, but our encoding \texttt{cw}
    significantly \textbf{improves} the performance of \textsf{ADDMC}, making
    $\textsf{ADDMC}+\texttt{cw}$ comparable to many other combinations. The
    scatter plot on the left-hand side adds to this by showing that \texttt{cw}
    is \textbf{particularly promising on certain data sets} such as DQMR and
    Grid networks. The scatter plot on the right-hand side shows that
    \texttt{cw} is better than \texttt{sbk05} (i.e., the second-best encoding
    for \textsf{ADDMC}) on the \textbf{majority of instances}. Seeing how, e.g.,
    DQMR instances are trivial for $\textsf{ADDMC}+\texttt{cw}$ but hard for
    $\textsf{Ace}+\texttt{cd06}$, and vice versa for Mastermind instances, we
    conclude that the best-performing algorithm-encoding combination
    \textbf{depends} significantly \textbf{on} (as-of-yet unknown)
    \textbf{properties of the Bayesian networks}.
  }

  \block{See the Paper For\dots}{
    \begin{itemize}
    \item Theoretical results that establish both the limits and the
      capabilities of classical WMC.
    \item Details of the \texttt{cw} encoding for Bayesian networks.
    \item A more detailed comparison of Bayesian network encodings, including a
      parameterized asymptotic analysis of their growth.
    \end{itemize}
  }
